# Differential abundance analysis

## Load the data
```{r load_data_mag_filtdamr_diffe, comment="", message=FALSE, warning=FALSE}
load("data/data.Rdata")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample))) %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

sample_metadata <- sample_metadata %>%
  filter(sample %in% colnames(genome_counts_filt))
```
## Create Phyloseq object
```{r phyloseq_gut, comment="", message=FALSE, warning=FALSE}
# Reorder levels: set "unsel" as reference
sample_metadata$fox_behaviour <- factor(
  sample_metadata$fox_behaviour,
  levels = c("unsel", "aggr", "tame")  #here we set unsel as reference level
)

#Phyloseq object
count_phy <- genome_counts_filt %>%
  column_to_rownames(var="genome")%>%
  otu_table(., taxa_are_rows=T)

sample_info_tab_phy <- sample_metadata%>%
  column_to_rownames(var="sample")%>%
  sample_data()

TAX <- genome_metadata%>%
  column_to_rownames(var="genome")%>%
  select(1:7)%>%
  as.matrix()%>%
  tax_table()
tree <- phy_tree(genome_tree)

physeq_all = phyloseq(count_phy, TAX, sample_info_tab_phy, tree)
```
## Run ANCOM-BC2 (fox behaviour)
Analysis of Composition of Microbes with Bias Correction

```{r ancombc_behaviour, comment="", message=FALSE, warning=FALSE}
set.seed(1234) #set seed for reproducibility
ancom_mag_gut = ancombc2(data = physeq_all, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "fox_behaviour + gut_location",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "fox_behaviour", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = TRUE,  #to get all pairwise comparisons
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```

## Phylum comparisons
### Aggressive vs Unselected
```{r ancom_rand1_gut, comment="", message=FALSE, warning=FALSE}
taxonomy <- data.frame(physeq_all@tax_table) %>%
  rownames_to_column(., "taxon")%>%
  mutate_at(vars(order, phylum, family, genus), ~ str_replace(., "[dpcofgs]__", ""))

ancom_mag_gut_table <- ancom_mag_gut$res %>%
  dplyr::select(taxon, lfc_fox_behaviouraggr, p_fox_behaviouraggr) %>%
  filter( p_fox_behaviouraggr < 0.05) %>%
  dplyr::arrange( p_fox_behaviouraggr) %>%
  merge(., taxonomy, by="taxon") %>%
  rename(genome=taxon) %>% 
  mutate_at(vars(phylum, genus), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_fox_behaviouraggr)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% #add 80% alpha
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(taxonomy$phylum))  # use all phyla from full taxonomy
colnames(tax_table)[1] <- "phylum"

tax_color <- merge(tax_table, colors_alphabetic, by="phylum") %>%
  arrange(phylum) %>%
  dplyr::select(colors) %>%
  pull()
```

```{r ancombc_rand_plot_mag_aggr_vs_unsel, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancom_mag_gut_table %>%
      mutate(genome=factor(genome,levels=ancom_mag_gut_table$genome)) %>%
ggplot(aes(x=lfc_fox_behaviouraggr, y=forcats::fct_reorder(genome,lfc_fox_behaviouraggr), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_vline(xintercept=0) + 
#  coord_flip()+
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  labs(title = "Aggressive vs Unselected")+
  xlab("log2FoldChange (aggr vs unsel)") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))
```


### Tame vs Unselected
```{r ancombc_rand_plot_mag_tame_vs_unsel, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancom_mag_gut_table_tame_unsel <- ancom_mag_gut$res %>%
  dplyr::select(taxon, lfc_fox_behaviourtame, p_fox_behaviourtame) %>%
  filter(p_fox_behaviourtame < 0.05) %>%
  arrange(p_fox_behaviourtame) %>%
  merge(taxonomy, by="taxon") %>%
  rename(genome=taxon) %>%
  arrange(lfc_fox_behaviourtame)

ggplot(ancom_mag_gut_table_tame_unsel,
       aes(x=lfc_fox_behaviourtame,
           y=fct_reorder(genome, lfc_fox_behaviourtame),
           fill=phylum)) +
  geom_col() +
  geom_vline(xintercept=0) +
  scale_fill_manual(values=tax_color) +
  labs(x="log2 Fold Change (tame vs unsel)", y="Species",
       title = "Tame vs Unselected")+
  theme_bw()


```
```{r ancom_mag_results}
ancom_mag_results <- ancom_mag_gut$res 
```


## Genus

### Aggressive vs Unselected
```{r ancombc_rand_plot_mag_aggr_vs_unsel_genus, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ggplot(ancom_mag_gut_table,
       aes(x=lfc_fox_behaviouraggr,
           y=fct_reorder(genome, lfc_fox_behaviouraggr),
           fill=genus)) +
  geom_col() +
  geom_vline(xintercept=0) +
  labs(x="log2 Fold Change (aggr vs unsel)", y="Species",
       title = "Aggressive vs Unselected")+
  theme_bw()
```


### Tame vs Unselected
```{r ancombc_rand_plot_mag_tame_vs_unsel_genus, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ggplot(ancom_mag_gut_table_tame_unsel,
       aes(x=lfc_fox_behaviourtame,
           y=fct_reorder(genome, lfc_fox_behaviourtame),
           fill=genus)) +
  geom_col() +
  geom_vline(xintercept=0) +
  labs(x="log2 Fold Change (tame vs unsel)", y="Species",
       title = "Tame vs Unselected")+
  theme_bw()
```

## Run ANCOM-BC2 (gut location)
Analysis of Composition of Microbes with Bias Correction

```{r ancombc_location, comment="", message=FALSE, warning=FALSE}
set.seed(1234) #set seed for reproducibility
ancom_mag_gut_location = ancombc2(data = physeq_all, 
                  assay_name = "counts", 
                  tax_level = NULL,
                  fix_formula = "gut_location+ fox_behaviour",
                  p_adj_method = "holm", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "gut_location", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = TRUE,  #to get all pairwise comparisons
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  lme_control = lme4::lmerControl(),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)

```


```{r dileum vs fcolon}
ancom_mag_gut_table_fcolon_dileum <- ancom_mag_gut_location$res %>%
  dplyr::select(taxon, lfc_gut_locationF_colon, p_gut_locationF_colon, q_gut_locationF_colon) %>%
  filter(p_gut_locationF_colon < 0.05) %>%
  arrange(p_gut_locationF_colon) %>%
  merge(taxonomy, by="taxon") %>%
  rename(genome=taxon) %>%
  arrange(lfc_gut_locationF_colon)

ggplot(ancom_mag_gut_table_fcolon_dileum,
       aes(x=lfc_gut_locationF_colon,
           y=fct_reorder(genome, lfc_gut_locationF_colon),
           fill=phylum)) +
  geom_col() +
  geom_vline(xintercept=0) +
  scale_fill_manual(values=tax_color) +
  labs(x="log2 Fold Change (D_ileum vs F_colon)", y="Species",
       title = "D_ileum vs F_colon")+
  theme_bw()
```
```{r dileum vs kileum}
ancom_mag_gut_table_kileum_dileum <- ancom_mag_gut_location$res %>%
  dplyr::select(taxon, lfc_gut_locationK_ileum, p_gut_locationK_ileum, q_gut_locationK_ileum) %>%
  filter(p_gut_locationK_ileum < 0.05) %>%
  arrange(p_gut_locationK_ileum) %>%
  merge(taxonomy, by="taxon") %>%
  rename(genome=taxon) %>%
  arrange(lfc_gut_locationK_ileum)

ggplot(ancom_mag_gut_table_kileum_dileum,
       aes(x=lfc_gut_locationK_ileum,
           y=fct_reorder(genome, lfc_gut_locationK_ileum),
           fill=phylum)) +
  geom_col() +
  geom_vline(xintercept=0) +
  scale_fill_manual(values=tax_color) +
  labs(x="log2 Fold Change (D_ileum vs K_ileum)", y="Species",
       title = "D_ileum vs K_ileum")+
  theme_bw()
```


## Volcano plots

```{r ancom_rand_volcano_mag_gut, comment="", message=FALSE, warning=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

#pdf("figures/different_species_StrucZero_new_violin.pdf",width=12, height=6)
ancom_mag_gut$res %>%
  dplyr::select(genome=taxon, lfc_fox_behaviouraggr, p_fox_behaviouraggr) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  mutate(phylum = ifelse(p_fox_behaviouraggr < 0.05, phylum, NA)) %>%
  ggplot(., aes(x = lfc_fox_behaviouraggr, y = -log(p_fox_behaviouraggr), color = phylum)) +
  geom_point(size=3, show.legend = FALSE) +
  #xlim(c(-10,4)) +
  scale_color_manual(values = phylum_colors) +
  labs(color = "Significant phylum", x = "Log-fold difference (aggr vs unsel)", y = "-log(p-value)") +
  theme_classic()
#dev.off()
```
