---
title: "AlberdiLab | Silver fox metagenomics"
subtitle: MSc project
author:
  - Irene Martínez, M Thomas P Gilbert, Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/silverfox_metagenomics
description: |
  Data analysis code for the study of Belyayev silverfox metagenomes
link-citations: yes
github-repo: alberdilab/silverfox_metagenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across ponds included in a restoration plan.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/silverfox_metagenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)
library(janitor)
library(broom)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)
library(UpSetR)
library(viridis)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(ANCOMBC)
library(lme4)
library(nlme)
library(pairwiseAdonis)
library(emmeans)
library(pheatmap)
library(rstatix)
library(uwot)
```

<!--chapter:end:index.Rmd-->

# Data preparation

## Load data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
sample_metadata <- read_csv("data/sample_metadata.csv") %>%
  rename(sample =1, 
         gut_location = 2,
         fox_behaviour = 3, 
         fox_sex = 4,
         sample_type = 5, 
         year = 6)
```


### Read counts

```{r load_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_tsv("data/counts.tsv") %>%
    rename(genome=1)%>%
    select(any_of(c("genome", sample_metadata$sample)))

```


### Filter read counts

```{r filter_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts_filt <- read_counts %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  select(one_of(c("genome",sample_metadata$sample)))

```


### Genome info

```{r load_genome_info, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_info <- read_csv("data/genomeInfo.csv") %>%
   mutate(genome = gsub(".fa$", "", genome)) %>%
   semi_join(., read_counts, by = "genome") %>%
   arrange(match(genome,read_counts$genome))
```

### Genome taxonomy

```{r load_genome_taxonomy, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_taxonomy <- read_tsv("data/genome_taxonomy.tsv") %>%
  rename(genome = 1)

genome_taxonomy_expanded <- genome_taxonomy %>% 
  separate(classification,
           into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
           sep = ";",
           fill = "right", remove = FALSE) %>%
  mutate(across(domain:species, ~ str_replace(.x, "^[a-z]__", "")))%>%
  mutate(phylum = case_when(
    phylum == "Actinobacteriota" ~ "Actinomycetota",
    phylum == "Firmicutes" ~ "Bacillota",
    phylum == "Firmicutes_A" ~ "Bacillota_A",
    phylum == "Firmicutes_C" ~ "Bacillota_C",
    phylum == "Cyanobacteria" ~ "Cyanobacteriota",
    phylum == "Proteobacteria" ~ "Pseudomonadota",
    TRUE ~ phylum))
```

### Genome metadata
```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_metadata <- genome_taxonomy_expanded %>%
  left_join(genome_info, by = "genome")
```


### Genome coverage

```{r load_bases, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_coverage <- read_tsv("data/bases.tsv") %>%
    rename(genome=1) %>%
    left_join(genome_info %>% select(genome, length), by = "genome") %>%
    mutate(across(where(is.numeric), ~ .x / length))
```


### Genome tree

```{r load_genome_tree, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("data/gtdbtk.backbone.bac120.classify.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```
### Genome annotations

```{r load_gene_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_annotations <- read_tsv("data/gene_annotations.tsv.xz") %>%
  mutate(genome = sub("\\^.*", "", gene)) %>%
  select(1, genome, everything())
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage

```{r filter_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  select(-length) %>% 
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]])) 
```


### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

### Distill annotations into GIFTs 

```{r distill_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol= 2, annotcol=c(6,7,8,9), verbosity = F)
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate(phylum=str_remove_all(phylum, "p__")) %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    dplyr::select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

location_colors <- c('#3D5C61','#41B6C0','#90C8C5','#E5D388','#BFA366','#6E5244')

origin_colors <- c("#bd70ae","#949293")
```

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_annotations,
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     genome_gifts, 
     phylum_colors,
     location_colors,
     origin_colors,
     file = "data/data.Rdata")
```


<!--chapter:end:01_data_preparation.Rmd-->

# MAG catalogue

```{r load_data_mag, message=FALSE, warning=FALSE, echo=FALSE}
load("data/data.Rdata")
```
## Genome phylogeny

Circular tree with genome size and completeness
```{r genome_phylogeny_completeness, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")


# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.5)


# Add phylum ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.05, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors) +
  labs(fill="Phylum")
        #theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=genome_metadata,
                geom=geom_bar,
                mapping = aes(x=completeness, y=genome, fill=contamination),
                offset = 0.24,
                pwidth = 0.1,
                orientation="y",
              stat="identity")+
  labs(fill="Contamination")

# Add genome-size ring
circular_tree <-  circular_tree + new_scale_fill()

circular_tree <-  circular_tree +
  geom_fruit(
    data=genome_metadata,
    geom=geom_bar,
    mapping = aes(x=length, y=genome),
    fill = "#1e6e55",
    offset = 0.05,
    orientation="y",
    stat="identity")


# Add text
circular_tree <-  circular_tree +
        annotate('text', x=2.15, y=0, label='            Phylum', family='arial', size=3.5) +
        annotate('text', x=2.6, y=0, label='                         Genome quality', family='arial', size=3.5) +
        annotate('text', x=2.8, y=0, label='                     Genome size', family='arial', size=3.5) 

#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```



Circular tree with genome presence/absence in different sample locations and fox behaviour
```{r genome_phylogeny, message=FALSE, warning=FALSE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate(phylum=str_remove_all(phylum, "p__")) %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    dplyr::select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate prevalence data
prevalence_data <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  group_by(genome,gut_location) %>% 
  summarise(presence=ifelse(sum(abundance)>0,1,0)) %>% 
  group_by(genome) %>%
  summarise(prevalence=sum(presence))

# Generate F_colon heatmap
f_colon_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(gut_location =="F_colon") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate M_colon heatmap
m_colon_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(gut_location =="M_colon") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate D_ileum heatmap
d_ileum_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(gut_location =="D_ileum") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate K_ileum heatmap
k_ileum_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(gut_location =="K_ileum") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate Gut content heatmap
gut_content_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(sample_type =="Gut content") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")


# Generate Gut tissue heatmap
gut_tissue_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(sample_type =="Gut tissue") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate tame heatmap
tame_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(fox_behaviour=="tame") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")

# Generate aggressive heatmap
aggressive_heatmap <- genome_counts_filt  %>%
  pivot_longer(!genome,names_to="sample",values_to="abundance") %>% 
  left_join(sample_metadata,by="sample") %>% 
  filter(fox_behaviour=="aggr") %>% 
  group_by(genome) %>% 
  summarise(presence=ifelse(sum(abundance)>0,"present","absent")) %>% 
  column_to_rownames(var="genome")
```

```{r genome_phylogeny_plot, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.3)

# Add phylum ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add F_colon ring
circular_tree <- gheatmap(circular_tree, f_colon_heatmap, offset=0.2, width=0.05, colnames=FALSE) +
        scale_fill_manual(values = c("absent" = "#ffffff", "present" = "#3D5C61")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add M_colon ring
circular_tree <- gheatmap(circular_tree, m_colon_heatmap, offset=0.3, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("#ffffff","#41B6C0")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add D_ileum ring
circular_tree <- gheatmap(circular_tree, d_ileum_heatmap, offset=0.4, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("#ffffff","#90C8C5")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add K_ileum ring
circular_tree <- gheatmap(circular_tree, k_ileum_heatmap, offset=0.5, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("absent" =  "#ffffff", "present" = "#E5D388")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add Gut content ring
circular_tree <- gheatmap(circular_tree, gut_content_heatmap, offset=0.7, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("absent" ="#ffffff","present" = "#BFA366")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add Gut tissue ring
circular_tree <- gheatmap(circular_tree, gut_tissue_heatmap, offset=0.8, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("#ffffff","#6E5244")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()



# Add prevalence ring
circular_tree <-  circular_tree +
        new_scale_fill() +
        scale_fill_manual(values = "#cccccc") +
        geom_fruit(
             data=prevalence_data,
             geom=geom_bar,
             mapping = aes(x=prevalence, y=genome),
                 offset = 0.5,
                 orientation="y",
         stat="identity")

# Add tame ring
circular_tree <- gheatmap(circular_tree, tame_heatmap, offset=1.4, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("#ffffff","#bd70ae")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add aggressive ring
circular_tree <- gheatmap(circular_tree, aggressive_heatmap, offset=1.5, width=0.05, colnames=FALSE) +
        scale_fill_manual(values=c("#ffffff","#949293")) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0)) +
        new_scale_fill()

# Add text
circular_tree <-  circular_tree +
        annotate('text', x=2.05, y= 0, label='Phylum', family='arial', size=3.5) +
        annotate('text', x=2.4, y=0, label='Gut location', family='arial', size=3.5) +
        annotate('text', x=3.1, y=0, label='Prevalence', family='arial', size=3.5) +
        annotate('text', x= 3.5,y=0, label='Fox behaviour', family='arial', size=3.5) 


#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```

### Taxonomy overview

```{r genome_taxonomy_bacteria}
tax_mag <-genome_metadata %>%
  group_by(phylum) %>%
  summarise(mag_n=n())

tax_mag %>%
  mutate(percetage_mag=round(mag_n*100/sum(mag_n), 2)) %>%
  arrange(-percetage_mag) %>%
  tt()
```

### Mag size (MB)

```{r mag size}
genome_metadata <- genome_metadata %>%
  mutate(corrected_size=100*length/completeness) %>%
  arrange(completeness)
```

```{r mag_size_all_mean, message=FALSE, warning=FALSE}
genome_metadata %>%
  summarise(Average_corrected_size=mean(corrected_size))
```

## Genome quality

```{r genome_quality}
tibble(Completeness=
         paste0(round(genome_metadata$completeness %>% mean(),2),
                "±",
                round(genome_metadata$completeness %>% sd(),2)),
       Contamination=
           paste0(round(genome_metadata$contamination %>% mean(),2),
                "±",
                round(genome_metadata$contamination %>% sd(),2))) %>%
  tt()
```


```{r genome_quality_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}

#Generate quality biplot
genome_biplot <- genome_metadata %>%
  dplyr::select(c(genome,phylum,completeness,contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=completeness,y=contamination,size=length,color=phylum)) +
              geom_point(alpha=0.7) +
                    xlim(c(70,100)) +
                    ylim(c(10,0)) +
                    scale_color_manual(values=phylum_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate contamination boxplot
genome_contamination <- genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_classic() +
                    theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(70,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_classic() +
                theme(legend.position = "none",
                    axis.line = element_blank(),
                    axis.title = element_blank(),
                    axis.text=element_blank(),
                    axis.ticks=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
#pdf("figures/completeness_contamination.pdf",width=10, height=5)
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))
#dev.off()
```


## Functional overview

#### Predicted genes
```{r predicted_stats, message=FALSE, warning=FALSE}
pred_genes <- genome_annotations %>%
  nrow()

cat(pred_genes)
```

#### Number of annotated genes and percentages
```{r annotation_stats, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
#How many genes have at least 1 annotation
genome_annota <- genome_annotations %>%
  filter(if_any(c(kegg, pfam, cazy), ~ !is.na(.))) %>%
  nrow()

cat(genome_annota)

#Percentage of predicted genes with at least 1 annotation
genome_annota*100/pred_genes
```

#### Number of KEGG annotatated genes and percentages
```{r kegg_stats_gut, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# KEGG annotation
kegg_annota <- genome_annotations %>%
  filter(!is.na(kegg)) %>%
  nrow()
cat(kegg_annota)

# KEGG annotation percentage
kegg_annota*100/genome_annota
```

#### Function tree
```{r function_heatmap, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db)

# Generate  basal tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(., size = 0.3) 

#Add phylum colors next to the tree tips
function_tree <- gheatmap(function_tree, phylum_heatmap, offset=0, width=0.1, colnames=FALSE) +
            scale_fill_manual(values=phylum_colors) +
            labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

#Add functions heatmap
function_tree <- gheatmap(function_tree, function_table, offset=0.5, width=3.5, colnames=FALSE) +
            vexpand(.08) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white") +
            labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

# Add completeness barplots
function_tree <- function_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\ncompleteness")

function_tree
```

## Functional ordination

PCoA with Euclidean distances:
```{r pcoa_ordination, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
gift_pcoa <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    vegdist(method="euclidean") %>%
    pcoa()

gift_pcoa_rel_eigen <- gift_pcoa$values$Relative_eig[1:10]


# Get genome positions
gift_pcoa_vectors <- gift_pcoa$vectors %>% #extract vectors
  as.data.frame() %>% 
  select(Axis.1,Axis.2) # keep the first 2 axes

gift_pcoa_eigenvalues <- gift_pcoa$values$Eigenvalues[c(1,2)]

gift_pcoa_gifts <- cov(genome_gifts, scale(gift_pcoa_vectors)) %*% diag((gift_pcoa_eigenvalues/(nrow(genome_gifts)-1))^(-0.5)) %>%
  as.data.frame() %>% 
  rename(Axis.1=1,Axis.2=2) %>% 
  rownames_to_column(var="label") %>% 
  #get function summary vectors
  mutate(func=substr(label,1,3)) %>% 
  group_by(func) %>% 
  summarise(Axis.1=mean(Axis.1),
            Axis.2=mean(Axis.2)) %>% 
  rename(label=func) %>% 
  filter(!label %in% c("S01","S02","S03"))
```

```{r pcoa_ordination_plot, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
set.seed(101)
scale <- 20 # scale for vector loadings
gift_pcoa_vectors %>% 
  rownames_to_column(var="genome") %>% 
  left_join(genome_metadata, by="genome") %>%
  ggplot() +
      #genome positions
      scale_color_manual(values=phylum_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color=phylum, size=length), 
                 alpha=0.9, shape=16) +
      #scale_color_manual(values=phylum_colors) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen[1]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA1")
            ) +
     scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen[2]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA2")
            ) +
    geom_label_repel(data = gift_pcoa_gifts,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
    xlim(-3,3) + 
    ylim(-2.5,2.5) +
    theme_minimal() + 
    theme(legend.position = "none")
```

PCoA with Bray-Curtis (relative composition)
```{r pcoa_ordination_bray, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
gift_pcoa_bray <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    vegdist(method="bray") %>%
    pcoa()

gift_pcoa_rel_eigen_bray <- gift_pcoa_bray$values$Relative_eig[1:10]


# Get genome positions
gift_pcoa_vectors_bray <- gift_pcoa_bray$vectors %>% #extract vectors
  as.data.frame() %>% 
  select(Axis.1,Axis.2) # keep the first 2 axes

gift_pcoa_eigenvalues_bray <- gift_pcoa_bray$values$Eigenvalues[c(1,2)]

gift_pcoa_gifts_bray <- cov(genome_gifts, scale(gift_pcoa_vectors_bray)) %*% diag((gift_pcoa_eigenvalues_bray/(nrow(genome_gifts)-1))^(-0.5)) %>%
  as.data.frame() %>% 
  rename(Axis.1=1,Axis.2=2) %>% 
  rownames_to_column(var="label") %>% 
  #get function summary vectors
  mutate(func=substr(label,1,3)) %>% 
  group_by(func) %>% 
  summarise(Axis.1=mean(Axis.1),
            Axis.2=mean(Axis.2)) %>% 
  rename(label=func) %>% 
  filter(!label %in% c("S01","S02","S03"))

```

```{r pcoa_ordination_plot_bray, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
set.seed(101)
scale <- 20 # scale for vector loadings
gift_pcoa_vectors_bray %>% 
  rownames_to_column(var="genome") %>% 
  left_join(genome_metadata, by="genome") %>%
  ggplot() +
      #genome positions
      scale_color_manual(values=phylum_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color=phylum, size=length), 
                 alpha=0.9, shape=16) +
      #scale_color_manual(values=phylum_colors) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts_bray, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen_bray[1]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA1")
            ) +
     scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen_bray[2]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA2")
            ) +
    geom_label_repel(data = gift_pcoa_gifts_bray,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
   xlim(-0.8,0.8) + 
    ylim(-0.8,0.8) +
    theme_minimal() + 
    theme(legend.position = "none")
```


UMAP

```{r umap_parameter_grid, message=FALSE, fig.height=12, fig.width=15}
set.seed(1001)

# Define parameter grid
neighbors_values <- c(5, 15, 30)
min_dist_values <- c(0.1, 0.3, 0.5)

umap_plots <- list()

# Loop over combinations
for (nn in neighbors_values) {
  for (md in min_dist_values) {
    
    umap_result <- umap(function_table,
                        n_neighbors = nn,
                        min_dist = md,
                        metric = "euclidean")
    
    df_umap <- as.data.frame(umap_result) %>%
      mutate(genome = rownames(function_table)) %>%
      inner_join(genome_metadata, by="genome") %>%
      rename(UMAP1 = V1, UMAP2 = V2)
    
    p <- ggplot(df_umap, aes(x = UMAP1, y = UMAP2,
                             color = phylum, size = length)) +
      geom_point(alpha = 0.7, shape=16) +
      scale_color_manual(values = phylum_colors) +
      theme_minimal() +
      labs(title = paste0("UMAP (neighbors=", nn,
                          ", min_dist=", md, ")"),
           color="Phylum", size="Genome size") +
      guides(color = guide_legend(override.aes = list(size = 5)))
    
    p <- p + theme(legend.position = "none")
    umap_plots[[paste(nn, md, sep="_")]] <- p
    
  }
}
ggpubr::ggarrange(plotlist = umap_plots, ncol = 3, nrow = 3)


```




```{r function_ordination_tsne, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
# Generate the tSNE ordination
set.seed(1001)
tSNE_function <- Rtsne(X=function_table, dims = 2, perplexity = 10, check_duplicates = FALSE)

# Plot the ordination
function_ordination <- tSNE_function$Y %>%
                as.data.frame() %>%
                mutate(genome=rownames(function_table)) %>%
                inner_join(genome_metadata, by="genome") %>%
                rename(tSNE1="V1", tSNE2="V2") %>%
                select(genome,phylum,tSNE1,tSNE2, length) %>%
                ggplot(aes(x = tSNE1, y = tSNE2, color = phylum, size=length))+
                            geom_point(shape=16, alpha=0.7) +
                            scale_color_manual(values=phylum_colors) +
                            theme_minimal() +
                labs(color="Phylum", size="Genome size") +
                guides(color = guide_legend(override.aes = list(size = 5))) # enlarge Phylum dots in legend

function_ordination
```



```{r testing_perplexity}
set.seed(1001)

# Trying different perplexity values
perplexities <- c(5, 7, 8, 9, 10, 11, 12, 13, 14)

plots <- lapply(perplexities, function(p) {
  tsne_result <- Rtsne(
    X = function_table, 
    dims = 2, 
    perplexity = p, 
    check_duplicates = FALSE
  )
  
  tsne_df <- as.data.frame(tsne_result$Y) %>%
    mutate(genome = rownames(function_table)) %>%
    inner_join(genome_metadata, by = "genome") %>%
    rename(tSNE1 = "V1", tSNE2 = "V2")
  
  ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, color = phylum)) +
    geom_point(alpha = 0.7, size = 2) +
    scale_color_manual(values=phylum_colors)+
    theme_minimal(base_size = 5) +  # make text larger
    theme(legend.position = "none") +  # remove legend
    labs(title = paste("Perplexity =", p))
})


ggpubr::ggarrange(plotlist = plots, ncol = 3, nrow = 3) 

```

<!--chapter:end:02_mag_catalogue.Rmd-->

# Community composition

```{r load_data_community, echo=FALSE}
load("data/data.Rdata")
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot_location, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_grid(. ~ gut_location,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x =element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size=8),
          axis.title.y = element_text(size=12),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.border = element_rect(colour = "black", fill = NA),
          strip.background = element_rect(fill = "white", color = "black"),
          strip.text = element_text(size = 12, lineheight = 0.6)) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```


```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~ gut_location + fox_behaviour,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x =element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size=8),
          axis.title.y = element_text(size=12),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.border = element_rect(colour = "black", fill = NA),
          strip.background = element_rect(fill = "white", color = "black"),
          strip.text = element_text(size = 12, lineheight = 0.6)) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

```{r taxonomy_barplot_2, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~  fox_behaviour + gut_location,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x =element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size=8),
          axis.title.y = element_text(size=12),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.border = element_rect(colour = "black", fill = NA),
          strip.background = element_rect(fill = "white", color = "black"),
          strip.text = element_text(size = 12, lineheight = 0.6)) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

```{r taxonomy_barplot_3, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~  fox_behaviour + sample_type,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x =element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size=8),
          axis.title.y = element_text(size=12),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.border = element_rect(colour = "black", fill = NA),
          strip.background = element_rect(fill = "white", color = "black"),
          strip.text = element_text(size = 12, lineheight = 0.6)) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

```{r taxonomy_barplot_4, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
    facet_nested(. ~   sample_type+ fox_behaviour,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(axis.text.x = element_blank(),
          axis.ticks.x =element_blank(),
          axis.title.x = element_blank(),
          axis.text.y = element_text(size=8),
          axis.title.y = element_text(size=12),
          panel.background = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.border = element_rect(colour = "black", fill = NA),
          strip.background = element_rect(fill = "white", color = "black"),
          strip.text = element_text(size = 12, lineheight = 0.6)) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```


### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,fox_behaviour) %>%
  summarise(relabun=sum(count))

phylum_summary %>%
    group_by(phylum) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              tame_mean=mean(relabun[fox_behaviour =="tame"]*100, na.rm=T),
              tame_sd=sd(relabun[fox_behaviour =="tame"]*100, na.rm=T),
              aggressive_mean=mean(relabun[fox_behaviour =="aggr"]*100, na.rm=T),
              aggressive_sd=sd(relabun[fox_behaviour=="aggr"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           tame=str_c(round(tame_mean,2),"±",round(tame_sd,2)),
           aggresive=str_c(round(aggressive_mean,2),"±",round(aggressive_sd,2))) %>% 
    arrange(-total_mean) %>% 
    select(phylum,total,tame,aggresive) %>% 
    tt()
```

```{r taxonomy_boxplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun)) %>%
    arrange(-mean) %>%
    dplyr::select(phylum) %>%
    pull()

phylum_summary %>%
  filter(phylum %in% phylum_arrange) %>%
  mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
    filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum, fill=phylum)) +
  scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        geom_boxplot(alpha=0.2)+
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ gut_location)+ 
        theme_minimal() + 
        theme(legend.position="none",
              strip.text.x = element_text(size = 14, color="black",face="bold"),
              axis.text.x = element_text(vjust = 0.5, size = 10),
              axis.text.y = element_text(size = 12),
              axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
              axis.title = element_text(size = 14, face = "bold"),
              axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
              axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0))
              ) +
        labs(y="Phylum",x="Relative abundance")
```

```{r taxonomy_boxplot_phylum_behaviour, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_summary %>%
  filter(phylum %in% phylum_arrange) %>%
  mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
    filter(relabun > 0) %>%
  ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum, fill=phylum)) +
  scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        scale_fill_manual(values=phylum_colors[-8]) +
        geom_boxplot(alpha=0.2)+
        geom_jitter(alpha=0.5) + 
        facet_nested(. ~ fox_behaviour)+ 
        theme_minimal() + 
        theme(legend.position="none",
              strip.text.x = element_text(size = 14, color="black",face="bold"),
              axis.text.x = element_text(vjust = 0.5, size = 10),
              axis.text.y = element_text(size = 12),
              axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
              axis.title = element_text(size = 14, face = "bold"),
              axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
              axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0))
              ) +
        labs(y="Phylum",x="Relative abundance")
```




<!--chapter:end:03_community_composition.Rmd-->

